A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-04-10 docs/manuscript.tex
last edits for preprint publication
new pdf for team review; update packages, update results file, update bib file

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{power analysis, data simulation, sample size, generalized linear mixed model\newline\indent Word count: 7600}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models},
  pdfauthor={Timo K. Koch,1, 2, Florian Pargent,1, Anne-Kathrin Kleine1, Eva Lermer1,3, \& Susanne Gaube1,4},
  pdflang={en-EN},
  pdfkeywords={power analysis, data simulation, sample size, generalized linear mixed model},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models}
\author{Timo K. Koch\textsuperscript{$\dagger{}$,1, 2}, Florian Pargent\textsuperscript{$\dagger{}$,1}, Anne-Kathrin Kleine\textsuperscript{1}, Eva Lermer\textsuperscript{1,3}, \& Susanne Gaube\textsuperscript{1,4}}
\date{}

\shorttitle{Simulation-Based Power Analysis for GLMMs}

@@

The authors made the following contributions. Timo K. Koch: Conceptualization, Formal Analysis, Writing - Original Draft Preparation, Writing - Review \& Editing; Florian Pargent: Formal Analysis, Methodology, Visualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Anne-Kathrin Kleine: Writing - Review \& Editing; Eva Lermer: Resources; Susanne Gaube: Writing - Review \& Editing, Supervision. \textsuperscript{$\dagger{}$} Timo K. Koch and Florian Pargent contributed equally to this work.

@@

\abstract{%
When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis in Human-Computer Interaction (HCI) and other empirical research fields underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori power analyses for GLMMs, necessitating data simulation approaches. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.
}

@@

When planning experimental research, it is essential to determine an appropriate sample size to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data (Lakens, 2022b). However, the recent replication crisis in Human-Computer Interaction (HCI) and several other disciplines grounded on empirical research has illustrated many challenges surrounding the reproducibility and reliability of findings (Cockburn, Dragicevic, Besançon, \& Gutwin, 2020; Robertson \& Kaptein, 2016; Yarkoni, 2022). As a result, there is a growing need for more rigorous statistical methodology and the adoption of well-powered experimental designs. While software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct ``a priori'' (i.e., before data collection) power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework in order to determine the required sample size in their experiments. In the present work, we provide a tutorial consisting of a concrete example for tailored a priori power analyses using data simulations based on GLMMs.

\hypertarget{statistical-power}{%
\subsection{Statistical power}\label{statistical-power}}

@@

Conducting underpowered research has many negative consequences. First, relying on underpowered experiments may yield inconclusive (if researchers acknowledge the small evidential value of an underpowered study in the limitations section) or misleading (if low power is ignored by the researchers) results, hindering the accumulation of knowledge. Second, underpowered studies waste resources by consuming time, effort, and funding without delivering meaningful results.

\hypertarget{a-priori-power-analysis}{%
\subsection{A priori power analysis}\label{a-priori-power-analysis}}

A power analysis represents the act of calculating the statistical power for a given true effect and sample size.
When running a power analysis before data collection, the required sample size can be determined so that researchers find an assumed true effect with the desired statistical power.

Thereby, a priori power analysis offers a valuable contribution to the research process by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value.
Moreover, conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for analyzing the data and answering their research questions.
Also, when conducting a proper power analysis, researchers have to consider every aspect of the experimental design and will notice statistical or design challenges before starting with data collection.
Adding a solid sample size calculation to the research process can act as a safeguard for ensuring high-quality research.
Finally, many journals and funding agencies now require that a power analysis is included in study protocols and grant proposals, recognizing its significance in ensuring robust and meaningful findings.

For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for a priori power analysis is readily available (Champely et al., 2018; Faul, Erdfelder, Buchner, \& Lang, 2009).
However, these software packages are often not flexible enough to perform power analysis for complex designs.

@@

As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and grouping structures introduced by their study designs (Yarkoni, 2022). GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data in HCI and other empirical disciplines because they offer a flexible framework for handling data with outcome variables that are not normally distributed (e.g., categorical outcomes) while accounting for both fixed and random effects (Fahrmeir, Kneib, Lang, \& Marx, 2021; Kaptein, 2016).

GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of linear regression models that account for correlated data including hierarchical structures (Fahrmeir et al., 2021).
In this context, correlated data means that the value in the outcome variable for one observation may be related (i.e., more similar or less similar) to the value for another observation in a systematic way that is not already accounted for by the usual (fixed) predictor variables (e.g., age of participants). This correlation can arise for various reasons: Responses to some stimuli from some participants might be more similar because the same person was measured twice (repeated measurements), both participants come from the same neighborhood (clustering) or both participants responded to the same stimulus (stimulus effects).
Thus, modeling such correlations is especially important whenever the data has a clear structure, while the grouping variables can be hierarchically organized (e.g., students nested in schools, schools nested in districts) or not (e.g., students solve math exercises, but neither student sees all exercises).
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on all fixed and random effects).
They allow for the modeling of fixed effects, which capture the relationships between our usual predictors and the outcome, as well as random effects, which account for the different types of correlation structure and grouping effects exemplified above.
Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across groups.

As mentioned, GLMMs extend the LMM framework to accommodate non-normally distributed continuous and categorical outcome variables.
GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear combination of predictor variables to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the outcome variable.
As an example, think of an experiment with different design factors (e.g., picture, headline) impacting the likelihood of users clicking on an online advertisement. Here, participants' behavior is measured repeatedly (e.g., over several sessions). The click patterns of participants in one session are likely to be correlated with their previous sessions. Finally, the outcome variable is binary (click/no click) for each interaction, which follows a binomial distribution.

\hypertarget{power-analysis-for-glmms}{%
\subsection{Power analysis for GLMMs}\label{power-analysis-for-glmms}}

Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods (Murayama, Usami, \& Sakaki, 2022).
Formula-based methods rely on often complicated formulas that can be used to directly calculate power while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., how often the hypothesis test is significant for the simulated data).
Currently available formula-based software packages for power analysis often do not include GLMMs or are limited to very simple designs (Murayama et al., 2022; Westfall, Kenny, \& Judd, 2014), making it necessary to build data simulations tailored specifically to the study design.
A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models (Arend \& Schäfer, 2019; Brysbaert \& Stevens, 2018; DeBruine \& Barr, 2021; Kumle, Võ, \& Draschkow, 2021; Lafit et al., 2021; Zimmer, Henninger, \& Debelak, 2022).
However, most of these tutorials focus on linear mixed models (LMMs) and the most common designs (but see Kumle et al., 2021 for a tutorial that also covers more advanced settings). This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available (see the discussion in Kumle et al., 2021).
The necessary presumptions for simulation-based power analysis with GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the correlation structure. The distributional assumption specifies the distributional family for the outcome variable (when conditioned on all fixed and random effects). Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Interpreting these presumptions entails understanding the underlying presumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge Kumle et al. (2021).

Based on the need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a simulation-based a priori power analysis with GLMMs. Thereby, we aim to equip researchers with the tools needed to simulate data and determine appropriate sample sizes for their own research.

\hypertarget{the-present-case-study}{%
\section{The present case study}\label{the-present-case-study}}

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs using a case study based on a specific experimental study design from the area of human-AI (artificial intelligence) interaction research. All code in this manuscript and simulation results are available in the project's repository on the Open Science Framework (\url{https://osf.io/dhwf4/}).

\hypertarget{experimental-study-design}{%
\subsection{Experimental study design}\label{experimental-study-design}}

In the present case study, we simulate data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system will be evaluated.
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head computer tomography (CT) scans to assess the presence of a bleeding.
To support their decision-making, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80\% of cases) or incorrect (20\%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans and if (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible.
The goal of the present simulation-based power analysis is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

\hypertarget{the-lme4-package-in-r}{%
\subsection{The lme4 package in R}\label{the-lme4-package-in-r}}

However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore, in this tutorial, we simulate data for GLMMs from first principles (i.e., creating synthetic data step by step instead of using black box functions) to assist applied researchers in better understanding all model assumptions and then use lme4 to analyze the simulated datasets.\footnote{A less flexible alternative would be to use the simr package (Green \& MacLeod, 2016), which can be used to both simulate data and perform power analysis for models supported by the lme4 package.}

\hypertarget{our-specific-glmm}{%
\subsection{Our specific GLMM}\label{our-specific-glmm}}

In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables \(\mathbf{X}\) and random effects \(\mathbf{U}\), transformed by a link function \(g()\) is modeled as a linear combination \(\eta\) of the predictor variables \(\mathbf{X}\), the random effects \(\mathbf{U}\) and the model parameters \(\mathbf{\beta}\) (Fahrmeir et al., 2021): \[
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
\] Equivalently, the conditional expected value is modeled as the linear combination \(\eta\), transformed by the inverse link function \(g^{-1}()\): \[
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
\] If the dependent variable (i.e., diagnostic decision) \(Y\) is a binary variable with values \(0\) (i.e., inaccurate), or \(1\) (i.e., accurate), the conditional expected value is equivalent to the probability: \[
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
\] In our case study, \(P_{si}\) is the conditional probability that a subject \(s\) gives the correct response to item (i.e., CT scan) \(i\).

In such a setting, we model this probability as \[
P_{si} = inverse\_logit(\eta_{si})
\] with the inverse-logit link \(g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}\) or equivalently \[
logit(P_{si}) = \eta_{si}
\] with the logit link \(g(P_{si}) = logit(P_{si}) = ln (\frac{P_{si}}{1 - P_{si}})\).

>>>>

\makeatletter
\usepackage{etoolbox}
\patchcmd{\maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{power analysis, data simulation, sample size, generalized linear mixed model\newline\indent Word count: 7504}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models},
  pdfauthor={Timo K. Koch,1, 2, Florian Pargent,1, Anne-Kathrin Kleine1, Eva Lermer1,3, \& Susanne Gaube1,4},
  pdflang={en-EN},
  pdfkeywords={power analysis, data simulation, sample size, generalized linear mixed model},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models}
\author{Timo K. Koch\textsuperscript{$\dagger{}$,1, 2}, Florian Pargent\textsuperscript{$\dagger{}$,1}, Anne-Kathrin Kleine\textsuperscript{1}, Eva Lermer\textsuperscript{1,3}, \& Susanne Gaube\textsuperscript{1,4}}
\date{}

\shorttitle{Simulation-Based Sample Size Planning for GLMMs}

@@

The authors made the following contributions. Timo K. Koch: Conceptualization, Formal Analysis, Writing - Original Draft Preparation, Writing - Review \& Editing; Florian Pargent: Formal Analysis, Methodology, Visualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Anne-Kathrin Kleine: Formal Analysis, Writing - Review \& Editing; Eva Lermer: Conceptualization, Resources, Writing - Review \& Editing; Susanne Gaube: Conceptualization, Supervision, Writing - Review \& Editing. \textsuperscript{$\dagger{}$} Timo K. Koch and Florian Pargent contributed equally to this work.

@@

\abstract{%
When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, available methods and software for a priori power analyses for GLMMs are often limited to specific designs, while data simulation approaches offer more flexibility. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.
}

@@

When planning experimental research, it is essential to determine an appropriate sample size to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data (Lakens, 2022b). However, the recent replication crisis has illustrated many challenges surrounding the reproducibility and reliability of findings (Yarkoni, 2022). As a result, there is a growing need for more rigorous statistical methodology and the adoption of well-powered experimental designs. While easy-to-use software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct ``a priori'' (i.e., before data collection) sample size planning for more complex research designs using the flexible generalized linear mixed models (GLMM) framework in order to determine the required sample size in their experiments. In the present work, we provide a tutorial on performing tailored simulation based sample size planning for GLMMs. After introducing general concepts, we review existing software solutions in R and discuss under which circumstances tailored data simulations are necessary. Then we describe general steps and decisions that are necessary whenever performing tailored data simulation. Because the details of these steps vary considerably for each specific study, we finish with a case study from the field of human-AI (artificial intelligence) interaction research.

--\textgreater{}

\hypertarget{prior-knowledge}{%
\section{Prior Knowledge}\label{prior-knowledge}}

TODO: fertig ausformulieren

Simulation-based sample size planning for generalized linear models is a complex topic that touches on a variety of statistical concepts and programming skills.
Although we give some short introduction to clarify our terminology, to profit the most from our tutorial, our readers require prior knowledge in the following topics (or refer to our collection of background literature):

Readers should understand basic statistical concepts like hypothesis tests and their statistical power but also confidence intervals and their precision.
Some background knowledge in empirical estimands, statistical estimates, and causal inference is very useful but not absolutely necessary.

In addition, readers should have basic programming skills in R on how to simulate data and how to repeat these simulations in parallel to speed up computation.
While there are a myriad of options for data simulation in R beyond the basic functionality in the base and stats packages, in our tutorial we use functions from the tidyverse and the faux package.

Finally, readers should be familiar with regression modeling in general and GLMMs in particular.
In our tutorial, we simulate data by manually specifying the model equation of a GLMM that represents our assumed data generating process.
While it is not necessary to understand the technical details of how GLMMs are estimated, it is crucial to understand the structure of a basic GLMM (e.g., logistic regression with random intercepts) and how the model assumes that the values in the dependent variable are determined by the predictor variables and the random effects.

Literature:

\begin{itemize}
\tightlist
\item
  Lakens ebook (Lakens, 2022a) evtl. Andy Field für die Basics?
\item
  R for data science (Wickham, Çetinkaya-Rundel, \& Grolemund, 2023), Bruine article (DeBruine \& Barr, 2021), faux documentation (DeBruine, 2023) für die R skills
\item
  kumle power tutorial (Kumle, Võ, \& Draschkow, 2021) und irgendwelche paper (am besten aus AMPPS) zu GLMMs
\item
  what is your estimand? paper (Lundberg, Johnson, \& Stewart, 2021) für das estimand topic
\end{itemize}

\hypertarget{general-concepts}{%
\section{General concepts}\label{general-concepts}}

\hypertarget{planning-for-statistical-power-or-precision}{%
\subsection{Planning for statistical power or precision}\label{planning-for-statistical-power-or-precision}}

Conducting research with too small sample sizes can have many negative consequences.
First, experiments may yield inconclusive or misleading results, hindering the accumulation of knowledge.
Second, studies that are doomed to never finding a postulated effect due to small sample size waste resources by consuming time, effort, and funding without delivering meaningful results.
For such reasons, many journals and funding agencies now require that a justification of sample size is included in study protocols and grant proposals, recognizing its significance in ensuring robust and meaningful findings.
Although many pragmatic justifications are possible (Lakens, 2022b), the ideal is often considered to determine a suitable sample size a priori (i.e., before the study is conducted) based on some meaningful computation to assure that the study will be able to fulfill its purpose.
In this way, adding a solid sample size calculation to the research process can act as a safeguard for ensuring high-quality research.

--\textgreater{}

The majority of empirical studies in psychology applies hypothesis testing.
As a consequence, the dominant approach for sample size planning is based on power analysis (i.e., planing for power).
However, sample size planning can also be based on the precision of parameter estimates (i.e., planning for precision).

\hypertarget{planning-for-power}{%
\subsubsection{Planning for power}\label{planning-for-power}}

@@

When planning for power, a target is set for the statistical power of a hypothesis test of interest (a very common heuristic is 0.8).
Assuming some effect size of interest and a desired significance level, a minimum sample size can be determined that (on average) would guarantee reaching this target.

\hypertarget{planning-for-precision}{%
\subsubsection{Planning for precision}\label{planning-for-precision}}

Not all research questions are best answered using hypothesis testing.
Especially for exploratory studies when little previous research has been conducted, only estimating effects of interest is often more useful.
When no hypothesis tests are conducted when analysing the planned study, power analysis is not not available for sample size planning.
Nonetheless, sample size still has a crucial effect on how informative the planned study will be to learn about the effect of interest.
In this case, the expected width of a confidence interval is used as a target for sample size planning.
The values inside a confidence interval are commonly interpreted as plausible values for the quantity of interest the confidence interval is supposed to estimate.
More formally, a 95\%-confidence interval provides the smallest interval with the property that upon repeated sampling, 95\% of individual confidence intervals would include the true quantity of interest.
Thus, a narrow confidence interval (assuming the confidence level is kept at a high value) is more informative about the size of the true effect than a wide interval.
Apart from the desired confidence level, the width of a confidence interval depends strongly on the sample size.
Because bigger samples carry more information, they lead to smaller confidence intervals (assuming everything else stays the same).
When planning for precision, a target is set for the expected width of a confidence interval of interest.
Assuming some effect size of interest and a desired confidence level, a minimum sample size can be determined that (on average) would guarantee reaching this target.

@@

As study designs become more complex, psychological researchers require more sophisticated statistical models to capture the nuanced relationships and grouping structures introduced by their study designs (Yarkoni, 2022).
GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data because they offer a flexible framework for handling data with outcome variables that are not normally distributed (e.g., categorical outcomes) while accounting for both fixed and random effects (Fahrmeir, Kneib, Lang, \& Marx, 2021).

GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of linear regression models that account for correlated data including hierarchical structures (Fahrmeir et al., 2021).
In this context, correlated data means that the value in the outcome variable for one observation may be related (i.e., more similar or less similar) to the value of another observation in a systematic way that is not already accounted for by the usual (fixed) predictor variables (e.g., age of participants).
This correlation can arise for various reasons: Responses to some stimuli from some participants might be more similar because the same person was measured twice (repeated measurements), both participants come from the same neighborhood (clustering) or both participants responded to the same stimulus (stimulus effects).
Thus, modeling such correlations is especially important whenever the data has a clear structure, while the grouping variables can be hierarchically organized (e.g., students nested in schools, schools nested in districts) or not (e.g., students solve math exercises, but neither student sees all exercises).
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on all fixed and random effects).
They allow for the modeling of fixed effects, which capture the relationships between our usual predictors and the outcome, as well as random effects, which account for the different types of correlation structure and grouping effects exemplified above.
Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across groups.

As mentioned, GLMMs extend the LMM framework to accommodate non-normally distributed continuous and categorical outcome variables.
GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear combination of predictor variables to the expected value of the outcome variable.
The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the outcome variable.
As an example, think of an experiment with different design factors (e.g., picture, headline) impacting the likelihood of users clicking on an online advertisement.
Here, participants' behavior is measured repeatedly (e.g., over several sessions).
The click patterns of participants in one session are likely to be correlated with their previous sessions.
Finally, the outcome variable is binary (click/no click) for each session, which follows a binomial distribution.

\hypertarget{simulation-based-sample-size-planning-with-glmms}{%
\section{Simulation based sample size planning with GLMMs}\label{simulation-based-sample-size-planning-with-glmms}}

For simpler statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for a priori sample size planning is readily available (Champely et al., 2018; Faul, Erdfelder, Buchner, \& Lang, 2009; Kelley, Maxwell, \& Rausch, 2003; Kelley \& Rausch, 2006).
However, these software packages are often not flexible enough to perform sample size planning for complex designs.

Existing approaches for sample size planning with GLMMs have almost exlusively focused on planning for power.
Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods (Murayama, Usami, \& Sakaki, 2022).
Formula-based methods rely on often complicated formulas that can be used to directly calculate power while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., how often the hypothesis test is significant for the simulated data).
Currently available formula-based software packages for power analysis with multilevel models often do not include GLMMs or are limited to very simple designs (Murayama et al., 2022; Westfall, Kenny, \& Judd, 2014), making it necessary to build data simulations tailored specifically to the study design.
A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models (Arend \& Schäfer, 2019; Brysbaert \& Stevens, 2018; DeBruine \& Barr, 2021; Kumle et al., 2021; Lafit et al., 2021; Zimmer, Henninger, \& Debelak, 2022).
However, most of these tutorials focus on linear mixed models (LMMs) and the most common designs (but see Kumle et al., 2021 for a tutorial that also covers more advanced settings). This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available (see the discussion in Kumle et al., 2021).
The necessary presumptions for simulation-based power analysis with GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the correlation structure. The distributional assumption specifies the distributional family for the outcome variable (when conditioned on all fixed and random effects). Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Interpreting these presumptions entails understanding the underlying presumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge Kumle et al. (2021).

\begin{itemize}
\tightlist
\item
  Insert table on existing R packages for power analysis with GLMMs
\end{itemize}

\hypertarget{when-do-we-profit-from-tailored-data-simulation}{%
\section{When do we profit from tailored data simulation?}\label{when-do-we-profit-from-tailored-data-simulation}}

Performing tailored simulation based sample size planning is more complicated and time consuming than using existing software tools for sample size planning.
Thus, it is worth discussing under which circumstances we most profit from or are forced to perform tailored simulation based sample size planning, because existing software solutions do not meet our requirements.

\hypertarget{complex-study-designs}{%
\subsection{Complex study designs}\label{complex-study-designs}}

TODO: fertig ausformulieren

Real studies are often more complex than the simplified designs assumed by many user-friendly software packages for sample size planning.

\hypertarget{missing-data}{%
\subsubsection{Missing data}\label{missing-data}}

TODO: fertig ausformulieren

Missing data is a frequent issue in applied data analysis.
There can be various reasons for why data is missing.
For example data can be missing completely at random.
Alternatively, subjects might drop out or produce missing data as a result of some covariate also measured in the dataset.
A a different note, many experimental designs contain some conditions in which some values of the predictor variables are missing by design.
Missing by design conditions can make data analysis more complicated because predictors have to be coded in highly specific ways that prevent the estimated GLMM to become unidentified.
Whether missing data has an effect on the result of sample size planning depends on the characteristics of the missing data process.
However, except for very simple scenarios, deciding whether missing data can be safely ignored in the sample size planning is often difficult to decide on a theoretical basis.
A better strategy is to explicitly include the (assumed) process of missing data into the data simulation process.

\hypertarget{data-generating-processes-beyond-standard-glmms}{%
\subsubsection{Data generating processes beyond standard GLMMs}\label{data-generating-processes-beyond-standard-glmms}}

TODO: fertig ausformulieren

While GLMMs are a very flexible model class that can handle a large variety of outcome variables, psychological researchers are becoming increasingly aware that many psychological datasets might profit from even more sophisticated models.
Common examples are zero-inflated outcomes, censoring, and nonlinear predictor effects.

\begin{verbatim}
- possibility to extend beyond standard GLMMs (e.g. zero-inflated outcomes, censoring, nonlinear effects, …)
\end{verbatim}

\hypertarget{complex-statistical-hypotheses}{%
\subsection{Complex statistical hypotheses}\label{complex-statistical-hypotheses}}

TODO: fertig ausformulieren

The most common hypotheses tested in psychological research are of the type \(H_0: \beta = 0\), where \(\beta\) is some coefficient in a regression model.
However, many research questions in psychology actually require testing more complex statistical hypotheses.

\hypertarget{directed-hypotheses-and-testing-against-values-other-than-0}{%
\subsubsection{Directed hypotheses and testing against values other than 0}\label{directed-hypotheses-and-testing-against-values-other-than-0}}

TODO: fertig ausformulieren

In the new era of preregistration and registered reports, most research questions are tested with directed hypotheses, because good theories should at least postulate whether some psychological effect of interest is positive or negative.
Even better theories should be able to specify a smallest effect sizes of interest (SESOI) that must be exceeded if the psychological effect has some practical relevance in a specific application.
In combination, this might require a test like \(H_0: \beta \leq 0.1\).

\hypertarget{combined-hypotheses}{%
\subsubsection{Combined hypotheses}\label{combined-hypotheses}}

TODO: fertig ausformulieren

Elaborate research questions often require testing hypotheses that consist of a combination of model parameters, for example \(H_0: \beta_0 + \beta_1 \leq 0\).
If the research question consists only of a single hypothesis of this sort, it is often possible to reduce the hypothesis to a single regression coefficient by clever coding and/or centering of predictor variables.
However, interesting research questions often consist of combined hypotheses that consist of more than one separate statistical hypotheses.
For example, a combined null hypothesis \(H_0\) might consist of two single null hypotheses \(H_{01}: \beta_1 \leq 0\) and \(H_{02}: \beta_0 + \beta_1 \leq 0\).
For some research questions, the combined null hypothesis \(H_0\) would be rejected if both \(H_{01}\) \emph{AND} \(H_{02}\) are rejected.
For other research questions, the combined null hypothesis \(H_0\) would be rejected if \(H_{01}\) \emph{OR} \(H_{02}\) \emph{OR} both are rejected.
If the global hypothesis \(H_0\) is combined with \emph{OR}, the p-values of the single hypotheses must be corrected for multiple testing to avoid \(\alpha\)-inflation for the global hypothesis.
However, if the global hypothesis \(H_0\) is combined with \emph{AND}, a correction for multiple testing is not necessary but rather a mistake that unnecessarily reduces the power of the global hypothesis test.

None of the software packages for sample size planning in table X can handle combined hypotheses as discussed here.
In contrast, our case study will demonstrate how we can test directed combined hypotheses with tailored simulation based sample size planning.

\hypertarget{planning-for-precision-1}{%
\subsection{Planning for precision}\label{planning-for-precision-1}}

TODO: fertig ausformulieren

As discussed earlier, the two major frameworks for sample size planning are planning for power and planning for precision.
Although planning for precision has been increasingly discussed as a more suitable strategy for exploratory research, all available software packages for sample size planning with GLMMs are based on power analysis.

As long as one were interested only in confidence intervals of a single model parameter, it might be possible to adapt existing software packages to plan for precision.
However, we demonstrate in our case study that in realistic research projects with GLMMs, the contrasts of interests and their respective confidence intervals are usually situated on the scale of the (expected) outcome in contrast to the scale of the model parameters.
We will show that tailored simulation based sample size planning can easily handle the planning for precision scenario.
The only change in procedure is that instead of computing hypothesis tests for each simulated dataset and estimating statistical power across repetitions, confidence intervals are computed for each simulated dataset and the expected width is estimated.

\hypertarget{no-prior-studies-or-pilot-data}{%
\subsection{No prior studies or pilot data}\label{no-prior-studies-or-pilot-data}}

TODO: fertig ausformulieren
